Experiences Gamifying Code Navigation

### 2013-10-03 paper_id1.tex
updated data
submitted to reading group for review

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Applying gamification to developer practices}
\subtitle{Why developer habits die hard}
% 
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
As software development practices evolve, we face the continuous challenge of communicating new practices and tools to the community.  As developers, we both like to try new things and like to stick with the familiar.  In addition to training, discussing, and presenting software engineering practices and tools, we seek to communicate them through active two-way information embedded in the Integrated Development Environment (IDE).

We collect data on current practices and tools used by the developer while giving them feedback on how they can improve.

 With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.

Blaze helps understand developer practice and tool use over the long term and judges the lasting change a communication effort makes in everyday habits.  Embeding Blaze n the developers work environment as a Visual Studio extension provides the capture of usage data while giving feedback to users on their practices.  Blaze encourages everyone to use the new practices by assigning points to specific commands and tools .  Providing a light score ranking view should help us stay motivated as we compare our command usage to our peers.  A month-long longitudinal study conducted with pilot users shows how develipers react to the presented information.

Results from this demonstrative study show the approach is <blank> with a few developers <blank> of the study.  The effectiveness of this approach for communicating about practices is demonstrated through the continued use of the commands as we communicate them.  Evaluation of metrics for navigation ratio over time show that developer use the commands <blank> than they did during the benchmarking phase.  The result of the study shows that this method for communicating and promoting software engineering tools and practices <blank>.
\end{abstract}

@@

As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.

@@

Perhaps the persistence of their habit is why developers are hard to turn from their old practices.  They may need some coaching to continue the new practice following a given lesson.  To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback to retain the attention and interest of the developer.  \cite{Snipes2013Towards} This tool creates a competitive game that rewards the developer continuously for adopting the improved practice.  The more they change their habits, the higher they score in the game.

@@

During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We evaluated developer's acceptance of game mechanics in the tool and their desires for additional functionality.  While further data is required to assess the impact of the game, the initial developer acceptance encourages an expanded study.

@@

\section{Study and Game Design}

Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.  Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models. \cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.  The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.  The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.

\begin{enumerate}
\item Define the practice we wish to measure compliance with
\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
\item Develop the intervention method
\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
\item Select participants to deploy
\item Define a pre-study questionnaire if one is necessary to collect associated information
\item Define a post-study questionnaire if necessary to collect post study information
\item Communicate with the participants about the study
\item Execute the study and report results
\end{enumerate}

@@

@@

After selecting the practice to focus on, we needed to select an intervention method.  The term intervention, indicates that we are applying a technique to influence a change in the measured behavior of developers.  Intervention involves communication about the practice in a way that motivates adoption \cite{}.  Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.  Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.

In the initial wave, participants were informed of the purpose of the study what to expect from the Blaze tool.  The participants installed the tool in their Visual Studio environment.  The tool did not present any window for them to see and simply collected data in the background.

After a week's data collection marking the end of the first wave, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our study completion survey.

In the third wave developers received instant feedback from Blaze on the points they accumulated during the development session.  The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.  This final configuration continued until the end of the pilot.

@@

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.  Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
%Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.

\subsection{Pilot Survey}

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

@@

\section{Related Work}
%update and expand

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.

@@

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  This builds on this work by testing methods for increasing the use of structured navigation in developers everyday practice.

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

@@

@@

\subsubsection{Observations about navigation ratio improvement}

To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a day over the number of unstructured navigation events.

Structured navigation events included:

Unstructured navigation events included:

This metric provides a continuous variable allowing us to use statistical data of continuous data to check our hypotheses.  To create the evaluation for each wave, we took the navigation ratio for each developer in each wave.

%The following hypotheses interest us in the current study.  
%1. Developers using blaze use more structured navigation when they receive instant feedback about their practices
%2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices

%To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.   Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.  Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.  What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study.  Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in wave 2 or wave 3.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{navigationaverage.pdf}
	\caption{Average Navigation Ratio by Week by Developer}
	\label{fig:navigationaverage}
\end{mdframed}\end{figure}

\subsubsection{Observations about efficiency of structured navigation}

Structured navigation sequences are less frequent between edits than unstructured navigation.  To study navigation practices in relationship to edits, we grouped the data into sequences delimited by edits meaning each sequence ends with one or more edits to the source code.   We evaluated the number of navigation commands between edits including clicks on locations in the file, navigating between files, and scrolling the file in the window.

We found that in x\% of the edit sequences structured navigation was sued.  We compared the time between edits where the sequences involved structured navigation to those that involved unstructured navigation.  The data supports the developer's perception that unstructured navigation is faster than structured navigation particularly when search is used.  We did find that when navigating between class files, developers who use structured navigation reached their destination X\% faster/slower than when using unstructured navigation.

Search was instrumental for developers who used the search tool where occurrences including search were completed x\% faster or slower than navigating by unstructured commands.

%  Our post pilot survey attempts to shed some light on the reasons behind the inconsistent improvement in navigation ratio.    Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.  The variation for each individual can be seen in Figure \ref{} .

@@

On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.  Other half of the participants have some understanding on structured navigation use it some of the time. While analyzing the points they scored with Blaze tool, we observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table. 
%the following statement is not supported with data unless you have additional data?
Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.

Assessing the game elements of Blaze that influenced them, four out of six said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.  Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.  The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.  The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.  The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.

@@

Comparing the survey demographic results with game scores reveals that years of experience did not have a strong influence over scores.  All developers in the study had 7 or more years experience and one person with 8 years experience used structured navigation enough to obtain second place nearly matching a 10 year experienced colleague.

Those who rated their knowledge of structured navigation commands highly, did better in the points than those who rated their prior knowledge lower.  This could indicate that using structured navigation is somewhat of a habit that takes longer to change.  There was also feedback that people wanted more information presented to them by the tool on what commands to use, so they may not have closed the knowledge gap during the pilot.

Those who worked in code that they wrote tended to use less structured navigation having a lower navigation ratio.  Those who had recently learned the code they were working in had more consistent use of structured navigation and one of them used Sando the most of the pilot participants.  Familiarity with the code thus may influence the use of structured navigation since when you maintain a mental map of the code, it may be more intuitive for you to click your way to it rather than use a search tool.  This may be an opportunity for future study to determine whether some developers maintain a somewhat abstract map of the code that makes it easier for them to click to it rather than recall keywords for a search.  %Could this be similar to how people navigate to software commands  in a menu structure more easily than using command line interfaces.

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\balance
\bibliography{paper_id1} 
 % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%

>>>>

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{International Conference on Software Engineering}{2014 Hyderbad, India}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Experiences Gamifying Code Navigation}
\subtitle{Are development practices a habit?}
% 
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
As software development practices evolve, we face the continuous challenge of communicating new practices and tools to the community.  As developers, we both like to try new things and like to stick with the familiar.  In addition to training, discussing, and presenting software engineering practices and tools, we propose a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).

Blaze monitors developer practice and tool use over the long term and reports the lasting change a communication effort makes in everyday habits.  Embeding Blaze in the developer`s work environment as a Visual Studio extension provides the capture of developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
 With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.

Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, the intended improvements in navigation practices promoted by the pilot did not occur.  We conclude that more proactive interventions are necessary to move developers to change their code navigation habits.
\end{abstract}

@@

%As  a developer I'm focused on the problem I'm currently solving and driven by productivity and deadlines to the point that I don't have time to sharpen the saw identifying better practices.

@@

Perhaps the persistence of their habit is why developers are hard to turn from their old practices.  Developers may need some coaching to continue a new practice following an introduction.  To continuously coach the developer, we conceived of a tool in our prior work to provide constant feedback and motivate them to adopt a new practice.  \cite{Snipes2013Towards} This tool creates a competitive game that gives continous feedback to the developer continuously on adopting the improved practice.  The more they change their habits, the higher they score in the game.

@@

During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We evaluated developer's acceptance of game elements in the tool and their desires for additional functionality.  While further data is required to assess the impact of the tool, the initial developer acceptance encourages an expanded pilot.

The rest of the paper is organized as follows:  Section 2 reviews the preliminary survey-based assement, Section 3 Discusses the design considerations for developing the Blaze tool/game, Section 4 Discusses the design of the study and participant selection, Section 5 is Related work, Section 6 results from the pilot, Section 7 Threats to Validity, and Section 8 has our Conclusions.

@@

\section{Game Design}

%\begin{enumerate}
%\item Define the practice we wish to measure compliance with
%\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
%\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
%\item Develop the intervention method
%\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
%\item Select participants to deploy
%\item Define a pre-study questionnaire if one is necessary to collect associated information
%\item Define a post-study questionnaire if necessary to collect post study information
%\item Communicate with the participants about the study
%\item Execute the study and report results
%\end{enumerate}

The pilot evaluates the aspects of game implementation as well as the study of the results of their results. Here we discuss how the aspects of selecting a practice suitable for gamification, designing the game, and competitive element considerations.

@@

\section {Study Design}

Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot to reveal feedback to the developer in 3 stages of 1 week duration each.
% Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models. \cite{longitudinalbook}  
We employed the three stage model defining the first stage as a baseline capturing the current practices of the pilot group.  The second stage employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.  The third stage provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.

@@

Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.  As previously discussed we sought to provide Blaze as a means to influence developers to use structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week that they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

In the initial stage, participants were informed of the purpose of the study what to expect from the Blaze tool.  The participants installed the tool in their Visual Studio environment.  The tool did not present any window for them to see and simply collected data in the background.

After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download another tool for code search called Sando \cite{Shepherd2012Sando} and information about using that tool in a video.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our study completion survey.

In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.  The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.  This final configuration continued until the end of the pilot.

@@

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.  Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.
%Through this approach, we ensured that the developers are performing the task on a very natural environment with out bothering that they are being tracked.
\subsection{Pilot Survey}

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers` navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

@@

\section{Related Work}
%update and expand
\subsection{Monitoring Practice Studies }
Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.

@@

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Key conclusions are that developers are more more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  This builds on this work by testing methods for increasing the use of structured navigation in developers` everyday practice.

@@

\subsection{Gamification}
Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

@@

Blaze logs events from actions the developer takes or actions from Visual Studio itself.
 The event is basically a GUI event that the program manages with event handlers that are registered with the application to listen for the event.  Blaze becomes a global event handler listening for all events by registering for them in Visual Studio.

Blaze records key attributes along with each event in visual studio.  The log captures a time-stamp for each event normalized to GMT.  The name of the event provides a low-level classification.  The event type reflects an internal classification in Blaze.  An optional field for Artifact Reference captures data such as the file-name being edited and the currently selected line number.  Finally an anonymized unique identifier for the user of blaze allows us to investigate differences between developers.

Analysis of the log extracts several key information bits from this raw source.  First we classify the event names at a higher level into categories related activities such as Navigating, Reading, Editing, Building, Debugging,  Testing, and using a  known Tool.

The second information we extract is aggregating the events along the time-line into sessions.  A session is an abstract concept that groups multiple events into a sequence with a beginning and an end.  Of the many possible ways to define sessions, we chose to define sequences as all the events that transpire before and during a set of edits to a specific file.  When the developer edits a new file, all the events after the last edit to the previous file are part of the new sequence for the new file.  Although not at the method level, this is similar to the say Robillard et. al. generated sequence based on located methods for their study.\cite{wbsnipes:Robillard2004How}

\subsubsection{Observations About Developers` Habits}

We are interested in understanding whether developers are consistently using practices that make them inefficient.  One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.  If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.  Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.

To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.  Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.  The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.

While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.  Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\

\subsubsection{Observations About Drivers of Edit Session Duration}

We selected navigation as the subject of the pilot game because it was identified by Robillard et. al\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    Are we focusing on the right area of developers` activity in Visual Studio?

In order to test whether navigation is an important area to address, we construct a hypothesis that use of unstructured navigation is an important factor in developer productivity.  Edit session duration, we assert by opinion, is a factor related to developer productivity though there are many other factors we could consider.  With the data available from Blaze, we can explore the categories of events that correlate with the duration of edit sessions over the period of the pilot.

Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search that selects attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.    The process identified the top 5 out of 10 possible categories as follows:
\begin{enumerate}
\item Unstructured Navigation
\item Read
\item Build
\item Debug
\item Edit
\end{enumerate}

Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.  The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 646 degrees of freedom.  The model with only unstructured navigation had an R-squared of 0.52 showing a good portion of the variation is explained by that category.  In order to have a more acceptable explanation of edit duration, we conducted a multiple linear regression model with both Unstructured Navigation and Read categories. The Read category of events captures the activities the developer is doing to look through the code in the editor.  They are navigation commands that move incrementally like pressing the up/down arrow keys, or moving the scroll-bar an incremental amount.  Combining these two categories results in a linear model with a p-value <0.001 and an R-squared of 0.76, more acceptable for drawing conclusions.

Conducting an ANOVA analysis shown in Table \ref{editDurationANOVA} for this regression model, we can explore the contribution of Unstructured Navigation and Reading to the model.  The contribution of each variable can be determined by calculating the portion of the total variation that each variable explains.  Unstructured Navigation explains 53\% of the variation.  Out of the remaining variation found by subtracting the amount explained by Unstructured Navigation from the total, Read explains 49\%.

\begin{table}[!t]

\begin{mdframed}[linecolor=white]
\begin{tabular}{|l|r|r|r|r|r|}
\hline
                          &  Df  & SSQ & MeanSq &Fval.   & Pr(>F)    \\
\hline
Unstr. Nav.& 1 & 0.039& 0.03920 & 1383 & <0.001\\
\hline
Read                 & 1 & 0.017 & 0.01741 & 614 & <0.001\\
\hline
Residuals                & 645 & 0.018 & 0.00003   &  &  \\
\hline
Total		& 647 & 0.074 & 0.05664 & & \\
\hline
\end{tabular}
\caption{ANOVA for Edit Duration}
\label{editDurationANOVA}
\end{mdframed}\end{table}

After this analysis we cannot disprove that unstructured navigation is correlated with edit duration, thus it is an important factor in developer productivity.  Combining unstructured navigation with reading describes a larger portion of the overall edit duration and motivates us towards improvements in both navigation methods and code comprehension.

\subsubsection{Observations About navigation ratio improvement}

To evaluate whether developers improved their practices for structured navigation we established a metric Navigation ratio as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.

Structured navigation events included: 
\begin{itemize}
\item Navigate To (Ctrl+,) is a fuzzy search interface that locates symbols matching your search terms.
\item Go To Definition (F12) brings up the code where the selected symbol is defined. 
\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of a symbol's dependencies and uses. 
\item Class View (Ctrl+W, C) priovides a searchable browser of classes and class hierarchy
\item Find All References provides a list of refernces for an object
\item Navigate to Event Handler brings up the event handler from the XAML editor
\item View Class Diagram generates a class diagram for the project
\item View Obeject Browser is a search tool for objects in the solution
\end{itemize}

Unstructured navigation events included selecting a file in an explorer window or selecting the tab for a file, using arrow and page up/down keys to go up/down through a file, scrolling, clicking on a file element, and using any of the built-in find commands such as Find In Files or Quick Find.

During the pilot we staged the deployment of information feedback to developers so they would gradually improve their structured navigation practices.  Our hypothesis was that this instant feedback and information would result in an increase in the use of structured navigation commands in Visual Studio.

To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.   Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.  Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{navratioboxplot.pdf}
	\caption{Navigation Ratio by Week}
	\label{fig:navigationaverage}
\end{mdframed}\end{figure}

What we observed visually in Figure \ref{fig:navigationaverage} is that the structured navigation practice did not increase significantly from the beginning to the end of the study.  Overall, we did not see a significant effect towards increasing navigation ratio for the interventions in week 2 or week 3.   Weeks 4 and 5 that follow the formal period of study show the navigation ratio continues in a similar distribution pattern.

@@

Our post pilot survey attempts to shed some light on the reasons behind the consistency in navigation ratio over the pilot.

On the assessment of structured navigation knowledge, half of the participants replied that they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.  Other half of the participants have some understanding on structured navigation use it some of the time. Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their more frequently using colleagues.  While analyzing the points they scored with Blaze tool, we observed that the participants who mentioned that they have good knowledge in structured navigation are on top of the points table. 
%the following statement is not supported with data unless you have additional data?
Incidentally, while cross verifying their performance on the coding, they are better productive compared to the participants on the bottom of points table.
\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{ElementInfluenceChart.pdf}
	\caption{Influence of Feedback Elements}
	\label{fig:elementInfluence}
\end{mdframed}\end{figure}

The chart in figure \ref{fig:elementInfluence} shows how the individual feedback features influenced developers to think about using structured navigation.  Four out of six developers said the leaderboard influenced them "a lot" and all rated the leaderboard as having some influence.  Three out of six  said their individual point score influenced them "a lot" and all said the score had some influence.  The element with mixed ratings was a graphical indicator on whether they were improving over past use of structured navigation commands.  The indicator received the influencing "a lot" rating from one participant and a "not at all" influence rating from one participant with the rest having some influence.  The results indicate the more obvious and more clearly comparative information in the leaderboard was the strongest influence for developers.

@@

%Though not significant, there is a loose correlation between those who rated their prior knowledge of structured navigation commands highly and their point scores.   
There was also feedback that people wanted more information presented to them by the tool on what structured navigation commands to use, so they may not have closed the knowledge gap during the pilot.

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\balance
\bibliography{../mousetap} 
 % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
